---
title: "ANA 515 Assignment 2 - SPY ETF Data Analysis"
author: "Shanmugapriya Mohan Kumar"
date: "`r Sys.Date()`"
output:
  html_document:
    theme:
      bootswatch: minty
---


<style>
  h1.title {
    text-align: center;
  }
  h4.author {
    text-align: center;
  }
  h4.date {
    text-align: center;
  }
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, message = TRUE, warning = FALSE)
library(tidyverse)
library(knitr)
library(bslib)
library(readr)
library(rmarkdown)
library(dplyr)  # for data manipulation
library(lubridate)  # for date handling
```

### Dataset Description
 
The dataset measures the daily stock performance of the SPDR S&P 500 ETF (SPY). This dataset contains variables such as <br>
1.	Open – Open price of that trading day.<br>
2.	High - High price of that trading day.<br>
3.	Low - Low price of that trading day.<br>
4.	Close - Closing price of that trading day.<br>
5.	Volume – trading volume of that trading day.<br>
6.	Date - date of that trading day.<br>

This data set is used to analyze historical price movements, identify trends and patterns, trading strategies, predict future prices, and make investment decisions.<br>
This data is collected from stock exchanges or financial data provider services such as Yahoo Finance or Google Finance.

**Research Questions:**<br>
1.	What are the trends and patterns of stock price movements?<br>
2.	How was the performance of SPY over a specific period?<br>
3.	Can historical data be used to predict future prices?<br>

**File Format:** File type is CSV and delimiter is Comma (,). The CSV files can be opened with Microsoft Excel, Google sheets, Statistical software such as R or Python’s Pandas library.


``` {r load-data, echo=TRUE}
# Dataset URL
spy_data <- "https://raw.githubusercontent.com/ShanmugapriyaMohankumar/StockMarket/main/SPY.csv"

# Read the dataset from the URL
spy500_data <- read_csv(spy_data,show_col_types = FALSE) #read_csv is part of the readr package, which is included in the tidyverse collection of R packages

#display the first few rows of the dataframe
head(spy500_data)

#Number of rows in the original dataset
spy500_data_row_count <- nrow(spy500_data)

#Number of columns in the original dataset
spy500_data_column_count <- ncol(spy500_data)

```

``` {r clean-data, echo=TRUE}
# Display the actual column names
colnames(spy500_data)

# Clean the data
spy500_data_clean <- spy500_data %>%
  rename(
    Open_price = Open,
    High_price = High,
    Low_price = Low,
    Close_price = Close
    ) %>%
  mutate(
    Date = ymd(Date)  # Convert the date column to Date format
  ) %>%
  arrange(desc(Volume))  # Arrange the data by volume in descending order

# Display the first few rows of the cleaned dataframe to verify the data is cleaned correctly
head(spy500_data_clean)

# Filter the dataset to include only the columns of interest
spy500_data_filtered <- spy500_data_clean %>%
  select(Date, Open_price, High_price, Close_price, Volume)

# Display the first few rows of the filtered dataframe
head(spy500_data_filtered)

# Find the date with the highest volume
highest_volume <- spy500_data_clean %>%
  slice_max(Volume, n=1) 

# Extract the date and volume
highest_volume_date <- highest_volume$Date
highest_volume_value <- highest_volume$Volume
```

### Date with the Highest Trading Volume:
The date with the highest trading volume is `r highest_volume_date`, with a volume of `r highest_volume_value`.<br>

### Data Frame Information
This data set has `r spy500_data_row_count` rows and `r spy500_data_column_count` columns.<br>
The names of the columns and a brief description of each are in the table below:<br>
  
  
```{r Column-description-table, echo=TRUE}
# Create a dataframe for column descriptions
column_descriptions <- data.frame(
  Column_number = c("1", "2", "3", "4", "5", "6"),
  Column = c("Date", "Open", "High", "Low", "Close", "Volume"),
  Description = c(
    "The date of the trading day",
    "The opening price of the trading day",
    "The highest price of the trading day",
    "The lowest price of the trading day",
    "The closing price of the trading day",
    "The trading volume of the trading day"
  )
)

# Display the table
kable(column_descriptions, col.names = c("Column Number", "Column Name", "Description"))
```

```{r summary-statistics, echo=TRUE}
# Create a subset of the dataframe with selected columns
spy500_data_subset <- spy500_data_filtered %>%
  select(Open_price, High_price, Close_price)

# Calculate summary statistics
summary_statistics <- spy500_data_subset %>%
  summarise(
    Open_price_min = min(Open_price, na.rm = TRUE),
    Open_price_max = max(Open_price, na.rm = TRUE),
    Open_price_mean = mean(Open_price, na.rm = TRUE),
    Open_price_na_count = sum(is.na(Open_price)),
    High_price_min = min(High_price, na.rm = TRUE),
    High_price_max = max(High_price, na.rm = TRUE),
    High_price_mean = mean(High_price, na.rm = TRUE),
    High_price_na_count = sum(is.na(High_price)),
    Close_price_min = min(Close_price, na.rm = TRUE),
    Close_price_max = max(Close_price, na.rm = TRUE),
    Close_price_mean = mean(Close_price, na.rm = TRUE),
    Close_price_na_count = sum(is.na(Close_price))
  )

# Display the summary statistics in a table
# Create a dataframe for the summary statistics
summary_df <- data.frame(
  Statistic = c("Minimum", "Maximum", "Mean", "Missing Values"),
  Open = c(summary_statistics$Open_price_min, summary_statistics$Open_price_max, summary_statistics$Open_price_mean, summary_statistics$Open_price_na_count),
  High = c(summary_statistics$High_price_min, summary_statistics$High_price_max, summary_statistics$High_price_mean, summary_statistics$High_price_na_count),
  Close = c(summary_statistics$Close_price_min, summary_statistics$Close_price_max, summary_statistics$Close_price_mean, summary_statistics$Close_price_na_count)
)
```

### Summary Statistics - S&P 500 Data(past 1 year)
```{r summary-statistics-display-table, echo=TRUE}
# Display the table
kable(summary_df, col.names = c("Statistic", "Open Price", "High Price", "Close Price"))

```


